---
title: "Adapting your data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adapting your data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

```{r setup, message = FALSE}
library(eventviz)
```

`plot_patient_journey()`'s defaults (`case_col = "caseID"`,
`time_col = "timestamp"`, ...) match `example_journey`'s column names, but
your own data almost certainly uses different ones — and might not even be
in the right *shape* yet. This vignette covers both problems: matching
column names via a schema, and reshaping wide milestone data into the long
form eventviz expects.

## Column-name schemas

You can always pass column names individually:

```{r manual, eval = FALSE}
plot_patient_journey(
  my_data, case_id = "12345",
  time_col = "ts", case_col = "spell_id", act_type_col = "category",
  activity_col = "label", patient_col = "mrn"
)
```

For a dataset you'll plot repeatedly, bundle the mapping into an
`event_log_schema()` once and reuse it:

```{r schema-manual}
my_schema <- event_log_schema(
  time_col = "ts", case_col = "spell_id",
  act_type_col = "category", activity_col = "label", patient_col = "mrn"
)
my_schema
```

Pass it as `schema = my_schema`; any individual argument you *also* supply
still wins over the schema, and anything neither supplies falls through to
`plot_patient_journey()`'s own defaults.

### Autodetection

If your column names are already conventional (`timestamp`/`time`,
`case_id`/`spell_id`/`episode_id`, `act_type`/`event_type`,
`activity`/`label`, `patient_id`/`mrn`, ...), you don't have to name them
at all. `autodetect_schema()` matches column names against built-in
candidate lists — exact case-insensitive match first, then a fuzzy
edit-distance match — and resolves roles in a fixed order so one column is
never silently claimed by two roles:

```{r autodetect}
autodetect_schema(example_journey)
```

Notice `caseID` and `K_Number` were matched fuzzily even though they don't
exactly match the candidate lists (`case_id`, `patient_id`, ...). To use
autodetection inline rather than as a separate step, pass the literal
string `schema = "auto"` — this is the *only* way autodetection runs; a
manually-constructed `event_log_schema()` never triggers it, and omitting
`schema` entirely just uses the function's hardcoded defaults:

```{r autodetect-inline}
plot_patient_journey(example_journey, case_id = "SP-001", schema = "auto")
```

If two candidate columns tie for one role, or a required role can't be
resolved at all, `autodetect_schema()` aborts naming the problem rather
than guessing — at that point, fall back to an explicit
`event_log_schema()`.

## Reshaping wide milestone data

Some source systems export one row *per case* with a column per milestone
timestamp, rather than one row per event:

```{r wide-data}
wide <- data.frame(
  case_id        = c("A", "B", "C"),
  diagnosis      = c("Chest pain", "Fall", "Sepsis"),
  arrival_time   = as.POSIXct(c("2024-01-01 08:00", "2024-01-01 09:15", "2024-01-01 10:00"), tz = "UTC"),
  triage_time    = as.POSIXct(c("2024-01-01 08:10", "2024-01-01 09:20", "2024-01-01 10:05"), tz = "UTC"),
  ward_time      = as.POSIXct(c("2024-01-01 11:30", NA,                  "2024-01-01 13:00"), tz = "UTC"),
  discharge_time = as.POSIXct(c("2024-01-01 18:00", "2024-01-01 12:00",  "2024-01-02 09:00"), tz = "UTC")
)
wide
```

`pivot_events_longer()` reshapes this into the long form
`plot_patient_journey()` needs, in one call. Name which of the pivoted
columns represent physical location moves via `location_cols` (everything
else pivoted becomes a point event by default), and any non-pivoted
columns (like `diagnosis` here) pass through untouched:

```{r pivot}
long <- pivot_events_longer(
  wide,
  case_col      = "case_id",
  time_cols     = c("arrival_time", "triage_time", "ward_time", "discharge_time"),
  location_cols = c("arrival_time", "ward_time", "discharge_time")
)
long
```

Two things happened automatically:

* Case B's missing `ward_time` (`NA`) was dropped with a message, rather
  than becoming a bogus event at a missing timestamp — an `NA` milestone
  means "didn't happen for this case", not an error.
* Activity labels were derived from the column names by stripping a
  trailing timestamp suffix (`_time`, `_at`, `_date`, `_ts`, `_datetime`)
  and title-casing what's left: `"arrival_time"` became `"Arrival"`, not
  `"Arrival Time"`.

Override either behaviour with `act_type_map`/`activity_map` (named
vectors from a `time_cols` entry to your preferred value) if the
auto-derived labels aren't what you want. The result feeds straight into
`plot_patient_journey()`:

```{r pivot-plot}
plot_patient_journey(
  long, case_id = "A",
  case_col = "case_id", location_categories = "location_move",
  patient_col = NULL
)
```

## Next steps

* `vignette("linear-processes")` for processes with no physical locations
  at all (complaints, tickets, approval pipelines).
* `vignette("cohort-analysis")` for comparing several cases once your data
  is in the right shape.
