---
title: "Adapting your data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adapting your data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5,
  error = TRUE
)
```

```{r setup, message = FALSE}
library(eventviz)
```

Every `eventviz` plotting function needs exactly one thing you must always
declare yourself: `state_events`, the `act_type` value(s) that open a
state. There's no default, and no value-level guessing — if you omit it,
the error tells you what's actually in your data:

```{r discovery-error}
orders <- data.frame(
  case_id   = c("A-1", "A-1", "A-1"),
  timestamp = as.POSIXct("2026-01-01") + c(0, 3600, 7200),
  act_type  = c("status", "note", "status"),
  activity  = c("Raised", "Chased supplier", "Approved")
)

plot_case_timeline(orders, case_id = "A-1")
```

The listing is ordered by frequency, so the value you almost certainly
want is usually first:

```{r discovery-fixed}
plot_case_timeline(orders, case_id = "A-1", state_events = "status")
```

Beyond `state_events`, `plot_case_timeline()`'s column-name defaults
(`case_col = "case_id"`, `time_col = "timestamp"`, ...) match a lot of data
out of the box, but your own data almost certainly uses different names —
and might not even be in the right *shape* yet. This vignette covers both
problems: matching column names via a schema, and reshaping wide milestone
data into the long form eventviz expects.

## Column-name schemas

You can always pass column names individually:

```{r manual, eval = FALSE}
plot_case_timeline(
  my_data, case_id = "12345", state_events = "status",
  time_col = "ts", case_col = "record_id", act_type_col = "category",
  activity_col = "label"
)
```

For a dataset you'll plot repeatedly, bundle the mapping into an
`event_log_schema()` once and reuse it:

```{r schema-manual}
my_schema <- event_log_schema(
  time_col = "ts", case_col = "record_id",
  act_type_col = "category", activity_col = "label",
  state_events = "status"
)
my_schema
```

Pass it as `schema = my_schema`; any individual argument you *also* supply
still wins over the schema, and anything neither supplies falls through to
`plot_case_timeline()`'s own defaults.

### Autodetection

If your column names are already conventional (`timestamp`/`time`,
`case_id`/`record_id`/`episode_id`, `act_type`/`event_type`,
`activity`/`label`, ...), you don't have to name them at all.
`autodetect_schema()` matches column names against built-in candidate
lists — exact case-insensitive match first, then a fuzzy edit-distance
match — and resolves roles in a fixed order so one column is never
silently claimed by two roles:

```{r autodetect}
renamed <- data.frame(
  case_idx  = c("A-1", "A-1"),
  event_ts  = as.POSIXct("2026-01-01") + c(0, 3600),
  act_typ   = c("status", "note"),
  activty   = c("Raised", "Chased supplier")
)
autodetect_schema(renamed)
```

`case_idx`, `event_ts`, `act_typ`, and `activty` were all matched fuzzily
even though none exactly match the candidate lists. To use autodetection
inline rather than as a separate step, pass the literal string
`schema = "auto"` — this is the *only* way autodetection runs; a
manually-constructed `event_log_schema()` never triggers it, and omitting
`schema` entirely just uses the function's hardcoded defaults:

```{r autodetect-inline}
plot_case_timeline(example_journey,
                   state_events = c("location_move", "ed_location_move"),
                   schema = "auto")
```

If two candidate columns tie for one role, or a required role can't be
resolved at all, `autodetect_schema()` aborts naming the problem rather
than guessing — at that point, fall back to an explicit
`event_log_schema()`.

## Reshaping wide milestone data

Some source systems export one row *per case* with a column per milestone
timestamp, rather than one row per event:

```{r wide-data}
wide <- data.frame(
  case_id        = c("A", "B", "C"),
  category       = c("Hardware", "Software", "Access"),
  received_time  = as.POSIXct(c("2024-01-01 08:00", "2024-01-01 09:15", "2024-01-01 10:00"), tz = "UTC"),
  triage_time    = as.POSIXct(c("2024-01-01 08:10", "2024-01-01 09:20", "2024-01-01 10:05"), tz = "UTC"),
  review_time    = as.POSIXct(c("2024-01-01 11:30", NA,                  "2024-01-01 13:00"), tz = "UTC"),
  closed_time    = as.POSIXct(c("2024-01-01 18:00", "2024-01-01 12:00",  "2024-01-02 09:00"), tz = "UTC")
)
wide
```

`pivot_events_longer()` reshapes this into the long form
`plot_case_timeline()` needs, in one call. Name which of the pivoted
columns mark a state change via `state_cols` — they emit
`act_type = "state_change"`, which is exactly what you then pass as
`state_events` (everything else pivoted becomes a point event by
default), and any non-pivoted columns (like `category` here) pass through
untouched:

```{r pivot}
long <- pivot_events_longer(
  wide,
  case_col   = "case_id",
  time_cols  = c("received_time", "triage_time", "review_time", "closed_time"),
  state_cols = c("received_time", "review_time", "closed_time")
)
long
```

Two things happened automatically:

* Case B's missing `review_time` (`NA`) was dropped with a message, rather
  than becoming a bogus event at a missing timestamp — an `NA` milestone
  means "didn't happen for this case", not an error.
* Activity labels were derived from the column names by stripping a
  trailing timestamp suffix (`_time`, `_at`, `_date`, `_ts`, `_datetime`)
  and title-casing what's left: `"arrival_time"` became `"Arrival"`, not
  `"Arrival Time"`.

Override either behaviour with `act_type_map`/`activity_map` (named
vectors from a `time_cols` entry to your preferred value) if the
auto-derived labels aren't what you want. The pivot and the plot agree out
of the box — the result feeds straight into `plot_case_timeline()`:

```{r pivot-plot}
plot_case_timeline(
  long, case_id = "A",
  case_col = "case_id", state_events = "state_change"
)
```

## Next steps

* `vignette("linear-processes")` for processes with no physical locations
  at all (complaints, tickets, approval pipelines).
* `vignette("cohort-analysis")` for comparing several cases once your data
  is in the right shape.
