---
title: "Cohort analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

```{r setup, message = FALSE}
library(eventviz)
```

Everything so far has looked at one case at a time. This vignette covers
comparing many cases at once: a faceted small-multiples view, and
aggregate statistics (per-stage durations, breach rates against a target,
and a transition-flow diagram) across a whole cohort. It uses
`complaint_example`'s eight complaints throughout.

## Faceted comparison: `plot_journey_cohort()`

`plot_journey_cohort()` lays several cases out as one facet panel per case.
By default (`case_ids = NULL`) it plots every case in the data, up to a
`max_cases` guard (25, since faceting hundreds of spells is a hang, not a
plot — pass explicit `case_ids` to raise it):

```{r cohort-absolute, fig.height = 9}
plot_journey_cohort(
  complaint_example,
  location_categories = "stage_change", case_col = "complaint_id",
  patient_col = NULL, terminal_activities = "Formal letter sent",
  state_label = "Stage"
)
```

Each panel gets its own time axis by default (`align_start = FALSE`), so
absolute dates line up across panels — useful for spotting whether several
complaints were affected by the same event (a staffing gap, a system
outage). The same location keeps the same fill colour in every panel: the
palette is built once, over the union of stages across the whole cohort,
not independently per panel.

### Aligning start times

To compare *durations* rather than *dates*, rebase every case to elapsed
hours from its own first move with `align_start = TRUE`. All panels then
share one `+Nh` axis:

```{r cohort-aligned, fig.height = 9}
plot_journey_cohort(
  complaint_example,
  location_categories = "stage_change", case_col = "complaint_id",
  patient_col = NULL, terminal_activities = "Formal letter sent",
  state_label = "Stage", align_start = TRUE
)
```

Now it's immediately visible that `"CMP-03"`'s bar runs far longer than
the others — the stall in "Under review" this vignette keeps coming back
to.

## Aggregate statistics

Faceting is good for eyeballing a handful of cases; for a whole cohort,
`eventviz` also has numeric summarisers built on the same box-derivation
pipeline, so they describe exactly what the plots draw.

### Per-stage duration statistics

```{r stage-durations}
summarise_stage_durations(
  complaint_example,
  location_categories = "stage_change", case_col = "complaint_id",
  patient_col = NULL
)
```

Every duration statistic in the package respects `end_inferred`: a case's
*final* stage has no recorded exit, so its dwell is an imputed rendering
convenience, not data. By default it's excluded from the mean/median/
quantiles (`n_inferred_excluded` reports how many were dropped);
pass `include_inferred = TRUE` to fold them back in.

### Breach rate against a target

`summarise_breach_rate()` reports what fraction of cases exceed a target,
either across the whole spell (`scope = "spell"`) or within one named
stage (`scope = "<stage name>"`, e.g. an SLA on how long a complaint may
sit in "Under review"):

```{r breach}
breach <- summarise_breach_rate(
  complaint_example, target_hours = 24 * 7, scope = "Under review",
  location_categories = "stage_change", case_col = "complaint_id",
  patient_col = NULL
)
breach
attr(breach, "breach_rate")
```

### Transition flow

`summarise_transitions()` reduces the cohort to directed stage-to-stage
transitions with counts and mean/median dwell in the origin stage;
`plot_transition_summary()` draws them as a flow diagram, with edge width
encoding frequency and forward/backward transitions bowed opposite ways so
re-entries stay legible:

```{r transitions, fig.height = 3}
plot_transition_summary(
  complaint_example,
  location_categories = "stage_change", case_col = "complaint_id",
  patient_col = NULL
)
```

Every complaint here follows the same six stages in the same order, so the
diagram is a straight line — the re-entry handling matters more for
cohorts where cases can loop back or skip stages (a ticket reopened after
"Resolved", say).

## Next steps

* `vignette("linear-processes")` for the single-case staircase view these
  aggregates complement.
* `?plot_patient_journey` for `return_data = TRUE`'s single-case `summary`
  element, built on the same `summarise_journey_durations()` this vignette
  uses across a cohort.
