---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

```{r setup, message = FALSE}
library(eventviz)
```

This vignette walks through `eventviz`'s core function,
`plot_case_timeline()`, on the bundled `example_journey` dataset: a
synthetic patient case moving through Ambulance arrival, Emergency
Department, Acute Medical Unit, Discharge Lounge, and Discharged, with
clinical point events (observations, tests, doctor reviews) sprinkled
throughout.

## The shape of an event log

`plot_case_timeline()` expects one row per event, with (at minimum) a
case identifier, a timestamp, an event-type column marking which rows
open a new exclusive state, and an activity/label column:

```{r}
head(example_journey)
```

Here `act_type == "ed_location_move"` or `"location_move"` marks a move to
a new location; everything else (`obs`, `clerk_review`, `test_ordered`,
...) is an instantaneous point event that gets plotted on the timeline but
doesn't create a new box. You tell `eventviz` which values open a state
via `state_events` — there's no default, and no guessing: if you omit it,
the error lists exactly what's in your data.

## The default plot

```{r default}
plot_case_timeline(example_journey,
                   state_events = c("location_move", "ed_location_move"))
```

`example_journey` has only one case, so `case_id` didn't need to be
supplied — `eventviz` used the only one available and said so. Each
coloured box is a state the case occupied over an interval; each point is
an instantaneous event, positioned on the midline of whichever box it fell
inside. The legend splits state fills from event-type shapes/colours so
the two never share a hue by default (the colourblind-safe `"okabe"`
palette — see `?journey_palette` for the `"brewer"` alternative).

## Duration labels

```{r show-duration}
plot_case_timeline(example_journey,
                   state_events = c("location_move", "ed_location_move"),
                   show_duration = TRUE)
```

The formatted duration appears above each non-terminal box. If a box's end
time had to be *inferred* (the data feed stopped before recording the next
state event), its label carries a trailing `"+"` so the plot never
overclaims precision it doesn't have.

## Terminal states

Without more information, the final box in a case has no natural end —
`plot_case_timeline()` has to infer one (by default, extending to the
last recorded event). For a case that reaches a genuinely *terminal*
state — "Discharged", "Closed", "Formal letter sent" — that inference is
wrong: the case didn't spend hours "being Discharged". Tell the function
which activities are terminal and it renders them as an instantaneous
marker instead:

```{r terminal}
plot_case_timeline(
  example_journey, case_id = "SP-001",
  state_events = c("location_move", "ed_location_move"),
  terminal_activities = "Discharged"
)
```

If a case's data feed simply stops before reaching a terminal state, the
box's right edge is drawn open (a dashed edge with an italic
`"(ongoing)"` label) rather than silently inventing an end — see the
[linear processes](linear-processes.html) vignette for a worked example
with `support_ticket_example`'s still-open ticket.

## Reference lines

Overlay a target/threshold line — useful for service standards like a
4-hour ED target:

```{r reference-lines}
plot_case_timeline(
  example_journey, case_id = "SP-001",
  state_events = c("location_move", "ed_location_move"),
  reference_lines = data.frame(offset_hours = 4, label = "4h target")
)
```

`reference_lines` takes a data frame with one row per line: `offset_hours`
(numeric, hours from the case's first event) and `label`.

## Direct labels and event bucketing

For a presentation-ready plot, `label_boxes = TRUE` labels each box
directly at its centre, and `event_type_top_n` collapses a long tail of
rare event types into `"Other"` so the legend stays readable:

```{r combined}
plot_case_timeline(
  example_journey, case_id = "SP-001",
  state_events = c("location_move", "ed_location_move"),
  label_boxes      = TRUE,
  event_type_top_n = 3
)
```

## Getting the underlying data

Pass `return_data = TRUE` to get the derived tables alongside the plot —
useful for debugging, or for building your own custom visualisation on top
of the same box/event derivation:

```{r return-data}
res <- plot_case_timeline(
  example_journey, case_id = "SP-001",
  state_events = c("location_move", "ed_location_move"),
  return_data = TRUE
)
names(res)
res$summary
```

`summary` is one row per state stay (case, state, entry/exit, duration,
and whether the end was inferred) — the same shape
`summarise_case_durations()` returns across a whole cohort. See the
[cohort analysis](cohort-analysis.html) vignette for more on aggregating
across cases.

## Next steps

* Bringing your own data? See `vignette("adapting-your-data")` for schema
  autodetection and the wide-to-long pivot wrapper.
* Got a process with no physical locations (a complaint, a ticket, an
  approval pipeline)? See `vignette("linear-processes")`.
* Comparing many cases at once? See `vignette("cohort-analysis")`.
