---
title: "Linear processes: complaints and tickets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear processes: complaints and tickets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

```{r setup, message = FALSE}
library(eventviz)
```

Not every event log has a spatial component. A complaint moves through
fixed handling stages; a support ticket moves through fixed statuses; a
purchase order moves through fixed approval steps. None of these cases
occupy a physical *location* — but each occupies exactly one *state*
exclusively at a time, which is exactly the structure
`plot_case_timeline()`'s box derivation was built for. This vignette
uses the bundled `complaint_example` and `support_ticket_example` datasets
to show both ways of visualising a linear process, and when to reach for
each.

## The datasets

`complaint_example` has eight complaints moving through
Acknowledgement -> Triage -> Assigned -> Under review -> Senior review ->
Formal letter sent, recorded as `act_type = "stage_change"`:

```{r complaint-head}
head(complaint_example)
```

`support_ticket_example` is the same shape for a support-ticket lifecycle
(Open -> Assigned -> In Progress -> Waiting on Customer -> Resolved ->
Closed) — deliberately non-healthcare, to prove the package isn't tied to
any one domain.

## Band layout: "what happened when"

`plot_case_timeline()` works unmodified — `stage_change`/`status_change`
plays the role `location_move` usually plays via `state_events`, and
`state_label` relabels the fill legend from "State" to something that fits:

```{r band}
plot_case_timeline(
  complaint_example, case_id = "CMP-01",
  state_events = "stage_change", case_col = "complaint_id",
  terminal_activities = "Formal letter sent",
  state_label = "Stage"
)
```

This is the right view for "what happened, and when" — the point events
(`contact`, `escalation`, `evidence_received`) sit on the timeline exactly
like any other point event.

## Staircase layout: "where does the time go"

For a linear process, the more compelling question is usually *where the
time goes* — which stage is eating the SLA. `plot_stage_ladder()` puts
the state on the y-axis (first stage at the top) and draws one horizontal
segment per stage, so the case walks down-and-right like a Gantt chart:

```{r ladder}
plot_stage_ladder(
  complaint_example, case_id = "CMP-01",
  state_events = "stage_change", case_col = "complaint_id"
)
```

Compare a complaint that stalls for weeks in one stage — the staircase
makes the bottleneck immediately visible in a way the band layout's
uniform box heights don't:

```{r ladder-stalled}
plot_stage_ladder(
  complaint_example, case_id = "CMP-03",
  state_events = "stage_change", case_col = "complaint_id"
)
```

### Per-stage targets

Pass `stage_targets` — a named vector of stage name to allowed dwell in
hours — to render a light allowance band on each targeted stage's row,
with any dwell beyond it redrawn in firebrick:

```{r ladder-targets}
plot_stage_ladder(
  complaint_example, case_id = "CMP-03",
  state_events = "stage_change", case_col = "complaint_id",
  stage_targets = c("Under review" = 24 * 7)   # one week
)
```

## Still-open cases

`support_ticket_example`'s `"TCK-04"` never reaches `"Closed"` — the data
feed simply stops. Passing `terminal_activities` tells
`plot_case_timeline()` what "done" looks like, so it can tell the
difference between a completed case and one still in flight; a case that
hasn't reached a terminal state gets its final box drawn open (a dashed
edge with an italic `"(ongoing)"` label) instead of silently inventing an
end for it:

```{r ongoing}
plot_case_timeline(
  support_ticket_example, case_id = "TCK-04",
  state_events = "status_change", case_col = "ticket_id",
  terminal_activities = "Closed", state_label = "Status"
)
```

## Which view should I use?

* **Band layout** (`plot_case_timeline()`) when you want the point
  events on the same timeline as the state moves, or you're comparing this
  process against genuinely spatial ones in the same report.
* **Staircase** (`plot_stage_ladder()`) when the question is "where's the
  bottleneck" for one case, especially with `stage_targets` set.
* Comparing many cases in either layout at once (rather than one at a
  time)? See `vignette("cohort-analysis")` — the staircase itself stays
  single-case for now; overlaying several cases on one ladder is a
  documented future direction, not yet implemented.
